---
title: "Box and Violin Plots"
author: "Michael Albers"
date: '2022-09-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Overview

Make a box plot and a violin plot and describe which is more effective and why


## Data Set


The Dow Jones Industrial Average (DJIA) index is a univariate time series from January 1, 1980 to December 31, 2012. The data set was downloaded from https://vincentarelbundock.github.io/Rdatasets/csv/AER/DJIA8012.csv


## Load Data set 


```{r Load dataframe}
djia <- read.csv('data/DJIA_Category.csv')
```


```{r}
head(djia)
```



```{r}
djia$category <- factor(djia$category, levels = c("80s decade", "90s decade", "2000s decade"))
```

```{r}
str(djia)
```


```{r}
summary(djia)
```

## Create Box Plot


```{r Box plot}
ggplot(djia, aes(x=category, y=value, fill=category)) +
  geom_boxplot(width = 0.5) +
  scale_fill_manual(name = "Decade", values = c("2000s decade" = "#ED813E", "90s decade" = "#9F8CFF", "80s decade" = "#1FD655")) +
  ylab("Index Value") +
  xlab("") +
  ggtitle("Dow Jones Industrial Average (DJIA) Index across decades")
```

The median DJIA index is higher for 2000s decade than both the 80s and 90s decades.  There are also outliers in 2000s decade indicating a period where there was a sharp drop in the Dow index.


## Create Violin Plot


```{r Violin Plot}
ggplot(djia, aes(x=category, y=value, fill=category)) +
  geom_violin(trim = FALSE) + coord_flip() + 
  scale_fill_manual(name = "Decade", values = c("2000s decade" = "#ED813E", "90s decade" = "#9F8CFF", "80s decade" = "#1FD655")) +
  stat_summary(fun.y = median, geom = "point", size = 2.5, color="yellow") +
  ylab("Index Value") +
  xlab("") +
  ggtitle("Dow Jones Industrial Average (DJIA) Index across decades")
```





## References

Dow Jones Industrial Average (DJAI), Retrieved from https://vincentarelbundock.github.io/Rdatasets/doc/AER/DJIA8012.html

