---
title: "Donut and Gauge Charts"
author: "Michael Albers"
date: '2022-10-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
```


## Overview

- Create Donut and Gauge chart

- Discuss which is most effective and why?

- Academic paper comments


## Data Sets

Wind farms in the United States of America downloaded from https://chart-studio.plotly.com/~jordanpeterson/781.csv


## Read Data Set

```{r}
wind_farm_df <- read.csv('data/wind-turbine-coord.csv')
```

```{r}
head(wind_farm_df)
```


## Donut Chart


### Top 10 Wind Turbine Manufacturers


```{r}
turbine_mfg_df <- wind_farm_df %>% 
  group_by(turbine_mfg) %>% 
  summarise(counts = n()) %>% 
  arrange(desc(counts))
```

```{r}
top_10_mfg <- top_n(turbine_mfg_df, 10)
```


```{r}
top_10_mfg_prop <- top_10_mfg %>% 
  mutate(prop = round(counts*100/sum(counts), 1)) %>% 
  arrange(desc(prop))
```

```{r}
top_10_mfg_prop$turbine_mfg
```



```{r}
top_10_mfg_prop$turbine_mfg <- factor(top_10_mfg_prop$turbine_mfg, levels = c("GE", "Vestas", "Kenetech", "Siemens",
                                                                              "Mitsubishi", "Gamesa", "Suzlon", "NEG Micon",
                                                                              "Bonus",  "Clipper" ))
```



```{r}
p <- ggplot(top_10_mfg_prop, aes(x = 2, y = prop, fill = turbine_mfg)) +
  geom_col() + 
  coord_polar("y", start = 100) +
  ggtitle("Top 10 Wind Turbine Manufacturers")
```


```{r}
p + geom_text(aes(label = paste(prop, "%", sep = "")), col = "white", size = 3.5, fontface = "bold",
            position = position_stack(vjust = 0.5)) +
    theme_void() +
    xlim(.2, 2.5)  
```



## References


Wood, Michael Lee (2020). "Visualizing Change in Ordinal Measures: Religious Attendance in the United States (1972-2018)." *Socius* 6:2378023120906944.



## Coding References


Donut chart in ggplot2. Retrieved from: 
https://r-charts.com/part-whole/donut-chart-ggplot2/


Changing the order of the slices in a pie chart in R. Retrieved from: 
https://stackoverflow.com/questions/70761001/changing-the-order-of-the-slices-in-a-pie-chart-in-r




