---
title: "Donut and Gauge Charts"
author: "Michael Albers"
date: '2022-10-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(skimr)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RColorBrewer)
```



## Overview


- Create Donut and Gauge charts

- Academic paper comments


## Data Set

Wind farms in the United States of America downloaded from https://chart-studio.plotly.com/~jordanpeterson/781.csv


## Read Data Set

```{r}
wind_farm_df <- read.csv('data/wind-turbine-coord.csv')
```

```{r}
head(wind_farm_df)
```


## Donut Chart


### Top 10 Wind Turbine Manufacturers


```{r}
turbine_mfg_df <- wind_farm_df %>% 
  group_by(turbine_mfg) %>% 
  summarise(counts = n()) %>% 
  arrange(desc(counts))
```

```{r}
top_10_mfg <- top_n(turbine_mfg_df, 10)
```


```{r}
top_10_mfg_prop <- top_10_mfg %>% 
  mutate(prop = round(counts*100/sum(counts), 1)) %>% 
  arrange(desc(prop))
```

```{r}
top_10_mfg_prop$turbine_mfg
```



```{r}
top_10_mfg_prop$turbine_mfg <- factor(top_10_mfg_prop$turbine_mfg, levels = c("GE", "Vestas", "Kenetech", "Siemens",
                                                                              "Mitsubishi", "Gamesa", "Suzlon", "NEG Micon",
                                                                              "Bonus",  "Clipper" ))
```



```{r}
p <- ggplot(top_10_mfg_prop, aes(x = 2, y = prop, fill = turbine_mfg)) +
  geom_col() + 
  coord_polar("y", start = 100) +
  ggtitle("Top 10 Wind Turbine Manufacturers")
```


```{r donut chart}
p + geom_text(aes(label = paste(prop, "%", sep = "")), col = "white", size = 3.5, fontface = "bold",
            position = position_stack(vjust = 0.5)) +
    theme_void() +
    labs(fill = "Turbine Mfg") +
    xlim(.2, 2.5) +
    scale_fill_brewer(palette = "RdYlGn")
  
```


This Donut chart displays the top ten wind turbine manufacturers in the U.S.  The GE manufacturer is the largest 
producer of wind turbines with 41% share of the top ten manufacturers.


## Gauge Chart


```{r}
summary(turbine_mfg_df)
```


```{r}
skim(turbine_mfg_df)
```

```{r GE proportion}
ge_prop <- round((16025*100 / sum(turbine_mfg_df$counts)), 1)
```


```{r gg.gauge function}
gg.gauge <- function(pos,breaks=c(0,30,70,100)) {
  require(ggplot2)
  get.poly <- function(a,b,r1=0.5,r2=1.0) {
    th.start <- pi*(1-a/100)
    th.end   <- pi*(1-b/100)
    th       <- seq(th.start,th.end,length=100)
    x        <- c(r1*cos(th),rev(r2*cos(th)))
    y        <- c(r1*sin(th),rev(r2*sin(th)))
    return(data.frame(x,y))
  }
  ggplot()+ 
    geom_polygon(data=get.poly(breaks[1],breaks[2]),aes(x,y),fill="red")+
    geom_polygon(data=get.poly(breaks[2],breaks[3]),aes(x,y),fill="gold")+
    geom_polygon(data=get.poly(breaks[3],breaks[4]),aes(x,y),fill="forestgreen")+
    geom_polygon(data=get.poly(breaks[4],breaks[5]),aes(x,y),fill="blue")+
    geom_polygon(data=get.poly(breaks[5],breaks[6]),aes(x,y),fill="purple")+
    geom_polygon(data=get.poly(pos-1,pos+1,0.2),aes(x,y))+
    geom_text(data=as.data.frame(breaks), size=5, fontface="bold", vjust=0,
              aes(x=1.1*cos(pi*(1-breaks/100)),y=1.1*sin(pi*(1-breaks/100)),label=paste0(breaks,"%")))+
    annotate("text",x=0,y=0,label=pos,vjust=0,size=8,fontface="bold")+
    coord_fixed()+
    theme_bw()+
    theme(axis.text=element_blank(),
          axis.title=element_blank(),
          axis.ticks=element_blank(),
          panel.grid=element_blank(),
          panel.border=element_blank()) 
}
```

**source:**  https://stackoverflow.com/questions/24900903/how-to-draw-gauge-chart-in-r

```{r test gauge plot}
gg.gauge(ge_prop, breaks = c(0,25,55,75,85,100)) +
  ggtitle("GE Manufacturer") +
  annotate("text", x=-0.55, y = -.065, label = "GE is largest wind turbine manufacturer in U.S.")
```


The Gauge chart includes all 69 wind turbine manufacturers with GE being the largest producer with 37.6% total market share in the U.S.


## Readings


**Visualizing Change in Ordinal Measures: Religious Attendance in the United States (1972-2018)**

The Likert-type bar plots do a good job in communicating the change in frequency of people attending religious services. It has a vertical line at the 50% threshold showing that the Reagan years had an increase in attendance to religious services, but there has been a gradual decline since. This article talks about the advantage of this chart over line charts having multiple categories. As a general rule, consider using a Likert chart if more than three categories exist and time is one of the measurements.



## References


Wood, Michael Lee (2020). "Visualizing Change in Ordinal Measures: Religious Attendance in the United States (1972-2018)." *Socius* 6:2378023120906944.



## Coding References


Donut chart in ggplot2. Retrieved from: 
https://r-charts.com/part-whole/donut-chart-ggplot2/


Changing the order of the slices in a pie chart in R. Retrieved from: 
https://stackoverflow.com/questions/70761001/changing-the-order-of-the-slices-in-a-pie-chart-in-r




