---
title: "Pie Charts Data Visualization"
author: "Michael Albers"
date: '2022-09-11'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Produce two pie charts and compare and contrast both charts. Discuss advantages and limitations of pie charts


## Data Set

The data set is the listing of companies that comprise the Dow Jones Industrial Average (DJIA) index as of September 8, 2022.  The data was downloaded from https://www.cnbc.com/dow-30/



```{r tidyverse library, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(scales)
```


## Read Data

```{r load dataframe}
dow30_df <- read.csv("data/DOW_30_SEPT_08.csv")
```

```{r}
head(dow30_df)
```


```{r}
str(dow30_df)
```


## Data Pre-processing

```{r create new gain_loss_desc column}
dow30_df$GAIN_LOSS_DESC <- with(dow30_df, ifelse(GAIN_LOSS == 2, 'Un-changed',
                                                 ifelse(GAIN_LOSS == 0, 'Loss', 'Gain')))
```


```{r convert columns to factor data types}
dow30_df$GAIN_LOSS_DESC <- as.factor(dow30_df$GAIN_LOSS_DESC)
dow30_df$SECTOR <- as.factor(dow30_df$SECTOR)
```

```{r}
str(dow30_df)
```

```{r}
head(dow30_df)
```


```{r gain_loss dataframe}
gain_loss_df <- dow30_df %>% 
  group_by(GAIN_LOSS_DESC) %>% 
  summarise(counts = n())
```

```{r}
gain_loss_df
```

```{r create prop and y-pos columns}
gain_loss_df <- gain_loss_df %>% 
  arrange(desc(GAIN_LOSS_DESC)) %>% 
  mutate(prop = round(counts * 100/sum(counts), 1),
         lab_y_position = cumsum(prop) - 0.5*prop)
```

```{r}
gain_loss_df
```

```{r sector dataframe}
sector_df <- dow30_df %>% 
  group_by(SECTOR) %>% 
  summarise(counts = n())
```

```{r}
sector_df <- sector_df %>% arrange(counts)
```


```{r}
sector_df
```



## Visualizations


```{r Greens palette}
show_col(brewer_pal(palette = "Greens")(5))
```

```{r Reds palette}
show_col(brewer_pal(palette = "Reds")(5))
```


### Gain/Loss Pie Chart

```{r init pie_gain_loss}
pie_gain_loss <- ggplot(gain_loss_df, aes(x = "", y = prop, fill=GAIN_LOSS_DESC))
```


```{r format pie_gain_loss}
pie_gain_loss + 
  ggtitle("Dow Jones Industrial Average - September 8, 2022") +
  geom_bar(width = 1, stat = "identity", color="white") +
  scale_fill_manual(name = "Gain/Loss", values = c("#006D2C", "#DE2D26", "gray")) +
  geom_text(aes(y = lab_y_position, label = prop), color = "white") +
  coord_polar("y", start = 0) +
  theme_void()
```


This pie chart uses the green color to indicate the proportion of companies in the DJIA index that had price gains. Red indicates the proportion of companies with a drop in stock price.  The grey color is a neutral color indicating that 3.3% of the DJIA stocks were unchanged.


### Sector Pie Chart


```{r}
sector_percent <- round(100 * sector_df$counts / sum(sector_df$counts), 1)
```

```{r}
sectors <- as.character(sector_df$SECTOR)
```


```{r, fig.height=6, fig.width=9 }
pie(sector_df$counts, labels = sector_percent, 
    main = "Dow Jones Industrial Average - September 8, 2022",
    col = rainbow(length(sector_df$counts)))
legend("topright", sectors, cex=0.75, title = "Sector Proportions", fill=rainbow(length(sector_df$counts)))
```


The sector pie chart is less effective than the pie chart for Gain/Loss.  It has more than seven categories which makes it difficult to understand.  This chart also exhibits color overload because it uses more than three colors.  A bar chart would be a better choice for the DJIA (Dow Jones Industrial Average) sector proportions.


## Summary


- A disproportional pie chart can distort the reality and make it appear that one pie slice is bigger than it really is. A donut chart can be a good alternative since the largest proportion does not dominate the visual as much as a pie chart does. 


- Pie charts can promote a scarcity mindset for the category with the smallest proportion


- Pie charts should be used to convey proportions of two or three categories as shown in the first pie chart

