---
title: "Scatter Plots"
author: "Michael Albers"
date: '2022-09-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(janitor)
```



## Overview

- Compare and contrast two scatter plots




## Data Set

This data set includes housing prices based on certain factors like house area, bedrooms, furnished, nearness to main road. The data set was downloaded from: https://www.kaggle.com/datasets/yasserh/housing-prices-dataset


**Acknowledgement:**

Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.
Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.



## Load Data

```{r}
housing_df <- read.csv('data/Housing.csv')
```

```{r}
head(housing_df)
```



## Pre-process Data

```{r}
str(housing_df)
```


```{r}
colnames(housing_df) <- c('price', 'area', 'bedrooms', 'bathrooms', 'stories', 'main_road', 'guest_room', 'basement', 'hot_water_heating', 'air_conditioning', 'parking', 'pref_area', 'furnishing_status')
```


```{r}
housing_df$main_road <- as.factor(housing_df$main_road)
housing_df$guest_room <- as.factor(housing_df$guest_room)
housing_df$basement <- as.factor(housing_df$basement)
housing_df$hot_water_heating <- as.factor(housing_df$hot_water_heating)
housing_df$air_conditioning <- as.factor(housing_df$air_conditioning)
housing_df$pref_area <- as.factor(housing_df$pref_area)
housing_df$furnishing_status <- as.factor(housing_df$furnishing_status)
```


```{r}
housing_df <- remove_empty(housing_df, which = "rows")
```


```{r}
str(housing_df)
```


```{r}

```


## House Price and Area Scatter Plot

```{r}
ggplot(data = housing_df, mapping = aes(x = area, y = price)) +
  geom_jitter(width = 0.001*sd(housing_df$price, na.rm = TRUE), alpha=0.5, color='blue') +
  labs(x="Area", y="Price")
```

```{r}

```


```{r}

```




## Summary

Scatter plots are helpful for identifying patterns between two numeric features. It can be helpful for identifying predictive features for a target variable.  Scatter plots are most useful for creating Linear Regression models for identifying features and analyzing residuals of a Linear Regression model.




