---
title: "Scatter Plots"
author: "Michael Albers"
date: '2022-09-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(janitor)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(scales)
```



## Data Set

This data set includes housing prices based on certain factors like house area, bedrooms, furnished, nearness to main road. The data set was downloaded from: https://www.kaggle.com/datasets/yasserh/housing-prices-dataset



## Load Data

```{r}
housing_df <- read.csv('data/Housing.csv')
```

```{r}
head(housing_df)
```



## Pre-process Data


- Rename column headers

- Convert character columns to factor datatype

- Scale Price in thousands



```{r}
str(housing_df)
```


```{r}
colnames(housing_df) <- c('price', 'area', 'bedrooms', 'bathrooms', 'stories', 'main_road', 'guest_room', 'basement', 'hot_water_heating', 'air_conditioning', 'parking', 'pref_area', 'furnishing_status')
```


```{r}
housing_df$main_road <- as.factor(housing_df$main_road)
housing_df$guest_room <- as.factor(housing_df$guest_room)
housing_df$basement <- as.factor(housing_df$basement)
housing_df$hot_water_heating <- as.factor(housing_df$hot_water_heating)
housing_df$air_conditioning <- as.factor(housing_df$air_conditioning)
housing_df$pref_area <- as.factor(housing_df$pref_area)
housing_df$furnishing_status <- as.factor(housing_df$furnishing_status)
housing_df$bedrooms <- as.factor(housing_df$bedrooms)
```


```{r}
housing_df <- remove_empty(housing_df, which = "rows")
```


```{r}
housing_df$price_thousands <- housing_df$price / 1000
```


```{r}
str(housing_df)
```



## House Price and Area Scatter Plots


```{r}
ggplot(data = housing_df, mapping = aes(x = area, y = price_thousands)) +
  geom_jitter(width = 0.001*sd(housing_df$price, na.rm = TRUE), alpha=0.5, color='blue') +
  labs(x="Area", y="Price (Thousands)") +
  ggtitle("Housing Scatter Plot 1")
```



The first scatter plot shows a relationship between Area and Price.  However, the relationship appears to weaken as Area increases above 5,000 square feet.


### Color by Bedrooms


```{r}
ggplot(data = housing_df, mapping = aes(x = area, y = price_thousands, colour = bedrooms)) +
  geom_jitter(width = 0.001*sd(housing_df$price, na.rm = TRUE), alpha=0.65) +
  scale_colour_brewer(palette = "Set1") +
  labs(x="Area", y="Price (Thousands)", color = "Number of Bedrooms") + 
  ggtitle("Housing Scatter Plot 2")
```


This scatter plot uses a color channel to encode the Bedrooms attribute.  This scatter plot reveals a stronger relationship between house area and price for one to three-bedroom homes, as opposed to houses with five and six bedrooms which appear more random.


```{r fig.width=11.25}
ggplot(data = housing_df, mapping = aes(x = area, y = price_thousands, colour = bedrooms)) +
  geom_jitter(width = 0.001*sd(housing_df$price, na.rm = TRUE), alpha=0.65) +
  labs(x="Area", y="Price (Thousands)", color = "Number of Bedrooms") + 
  scale_colour_brewer(palette = "Set1") +
  ggtitle("Housing Scatter Plot 3") +
  facet_grid(cols = vars(bedrooms))
```


The final chart is grouping of scatter plots by the Bedrooms attribute.  It confirms that houses with two to four bedrooms show a stronger relationship between Area and Price.



## Summary

Scatter plots are helpful for identifying patterns between two numeric features. It can be helpful for identifying predictive features for a target variable.  Scatter plots are useful for creating Linear Regression models for identifying features and analyzing residuals of a Linear Regression model.




